A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-01-18 sections/algorithm.tex
some alg. texts

$$\begin{aligned}
\text{Type variables}\qquad&a, b\\
\text{Existential variables}\qquad&\al, \bt\\[3mm]
%\text{Types}\qquad&A', B', C' &::=&\quad 1 \mid a \mid \forall x. A' \mid A'\to B'\\
\text{Mono-types}\qquad&\tau &::=&\quad 1 \mid a \mid \tau_1\to \tau_2\\
%\text{Context}\qquad&\Psi &::=&\quad \cdot \mid \Psi, a \mid \Psi, x:A' \color{red} \\[3mm]
% \text{Declarative Judgements}\qquad&\jo &::=&\quad \cdot \mid \P\vdash A' \le B' \jc \jo\\
% &&&\quad \mid \P\vdash e\Leftarrow A' \jc \jo \mid \P\vdash e\To A' \jc \jo \mid \P\vdash A' \bullet e\TTo C' \jc \jo\\
% \noalign{\jimmy{\text{What do the markers do: $\G \vdash,a \vdash,b\vdash \j_1;\j_2;\j_3$ means $\G,a,b\vdash \j_1 \land \G,a\vdash \j_1,\j_2 \land \G\vdash \j_3$}}}\\[3mm]
\text{Algorithmic types}\qquad&A, B, C &::=&\quad 1 \mid a \mid \al \mid \forall a. A \mid A\to B\\
\text{Algorithmic judgement chain}\qquad&\jg &::=&\quad A \le B \mid e\Lto A \mid e\To_{a} \jg \mid \appInfAlg{A}{e}\\[3mm]
\text{Algorithmic worklist}\qquad&\Gm &::=&\quad \cdot \mid \Gm, a \mid \Gm, \al \mid \Gm, x: A \mid \Gm \vdash \jg\\[3mm]
\text{Terms}\qquad&e&::=&\quad x \mid () \mid \lam{e} \mid e_1~e_2 \mid (e:A)
\end{aligned}$$

@@

\paragraph{Judgment and judgment chain} ...
\paragraph{Worklist Judgments} Our algorithmic judgment has a non-standard form.
We combine the traditional contexts and judgment(s) into a single worklist.
The worklist collects variable bindings and judgments in order, which represents their scopings:
each variable is visible after it is declared.
For example, $a,x:a,x \Lto a$ is a valid worklist, but $\textcolor{red}{x} \Lto a, x:\textcolor{red}{a},a$ is not.

@@

\begin{figure}[t]
\framebox{$\Gm\vdash\jg$} Well-formed algorithmic judgment.
\begin{gather*}
\inferrule*[right=$\mathtt{wf{\le}}$]
{\Gm\vdash A \\ \Gm\vdash B}
{\Gm\vdash A\le B}
\qquad
\inferrule*[right=$\mathtt{wf{\Lto}}$]
{\Gm\vdash e \\ \Gm\vdash A}
{\Gm\vdash e \Lto A}
\\
\inferrule*[right=$\mathtt{wf{\To}}$]
{\Gm\vdash e \\ \Gm, a\vdash \jg}
{\Gm\vdash e \To_a \jg}
\qquad
\inferrule*[right=$\mathtt{wf{\TTo}}$]
{\Gm\vdash A \\ \Gm, a\vdash \jg \\ \Gm\vdash e}
{\Gm\vdash \appInfAlg{A}{e}}
\end{gather*}

\framebox{$\text{wf }\Gm$} Well-formed algorithmic worklist.
\begin{gather*}
\inferrule*[right=$\mathtt{wf\cdot}$]
{~}{\text{wf }\cdot}
\qquad
\inferrule*[right=$\mathtt{wf_a}$]
{\text{wf }\Gm}
{\text{wf }\Gm, a}
\qquad
\inferrule*[right=$\mathtt{wf_{\al}}$]
{\text{wf }\Gm}
{\text{wf }\Gm, \al}
\qquad
\inferrule*[right=$\mathtt{wf_{of}}$]
{\text{wf }\Gm \\ \Gm\vdash A}
{\text{wf }\Gm, x:A}
\qquad
\inferrule*[right=$\mathtt{wf_{\jg}}$]
{\text{wf }\Gm \\ \Gm\vdash\jg}
{\text{wf }\Gm \vdash\jg}
\end{gather*}
\caption{Algorithmic Well-Formedness}\label{fig:alg}
\end{figure}

@@

% TODO rules too long
\begin{figure}[t]
\framebox{$\Gm\rto \Gm'$} $\Gm$ reduces to $\Gm'$.
%\jimmy{Whenever there is a change on judgment count, marker should be properly added/removed. Decidability: all valid judgment chain should be reduced to nil.}
\begin{gather*}
\begin{aligned}
\Gm, a &\rto \Gm \qquad\qquad
\Gm, \al \rto \Gm \qquad\qquad
\Gm, x:A \rto \Gm
\\[2mm]
\Gm \vdash 1\le 1 &\rto \Gm\\
\Gm \vdash a\le a &\rto \Gm\\
\Gm \vdash \al\le \al &\rto \Gm
\\[2mm]
\Gm \vdash A_1\to A_2 \le B_1\to B_2 &\rto \Gm \vdash A_2 \le B_2 \vdash B_1\le A_1\\
\Gm \vdash \forall a. A\le B &\rto \Gm,\al \vdash [\al/a]A\le B \quad\text{\jimmy{No more existential variable "leaks"!}}\\
\Gm \vdash A\le \forall b. B &\rto \Gm,b \vdash A\le B \quad\text{\jimmy{And no more type variable "leaks"!}}
\\[2mm]
%\\
%\text{Let } \color{red}\Gm[\al \toto B // G_M] &:= \Gm_L, G_M, [B/\al]\Gm_R, \text{ when } \Gm[\al] = \Gm_L,\al,\Gm_R \ (G_M\text{ defaults to }\cdot)\\
\Gm[\al] \vdash \al \le A\to B &\rto [\al[1]\to\al[2]/\al] (\Gm[\al[1], \al[2]] \vdash \al[1]\to \al[2] \le A \to B)\\
 &\qquad \text{when }\al\notin FV(A)\cup FV(B)\\
\Gm[\al] \vdash A\to B \le \al &\rto [\al[1]\to \al[2]/\al] (\Gm[\al[1], \al[2]] \vdash A \to B \le \al[1]\to \al[2])\\
 &\qquad \text{when }\al\notin FV(A)\cup FV(B)
 \\[2mm]
\Gm[\al][\bt] \vdash \al \le \bt &\rto [\al/\bt](\Gm[\al][])\\
\Gm[\al][\bt] \vdash \bt \le \al &\rto [\al/\bt](\Gm[\al][])\\
\Gm[a][\bt] \vdash a \le \bt &\rto [a/\bt](\Gm[a][])\\
\Gm[a][\bt] \vdash \bt \le a &\rto [a/\bt](\Gm[a][])\\
\Gm[\bt] \vdash 1 \le \bt &\rto [1/\bt](\Gm[])\\
\Gm[\bt] \vdash \bt \le 1 &\rto [1/\bt](\Gm[])
\\[2mm]
\Gm \vdash e \Lto B &\rto \Gm \vdash e\To_a a\le B\\
\Gm \vdash () \Lto 1 &\rto \Gm\\
\Gm \vdash e\Lto \forall a. A &\rto \Gm,a \vdash e\Lto A\\
\Gm \vdash \lambda x. e \Lto A\to B &\rto \Gm, x:A  \vdash e \Lto B\\
\Gm[\al] \vdash \lambda x. e \Lto \al &\rto [\al[1]\to \al[2] / \al](\Gm[\al[1],\al[2]], x:\al[1] \vdash e \Lto \al[2]) \quad\text{\jimmy{Additional case}}
\\[2mm]
\Gm \vdash x\To_a \jg &\rto \Gm \vdash [A/a] \jg \qquad (x:A)\in \Gm\\
\Gm \vdash (e:A)\To_a \jg &\rto \Gm \vdash [A/a]\jg \vdash e \Lto A\\
\Gm \vdash ()\To_a \jg &\rto \Gm \vdash [1/a]\jg \qquad\jimmy{\downarrow\text{Precise control on exvars: to the $2^{nd}$ chain}}\\
\Gm \vdash \lambda x. e \To_a \jg &\rto
	\Gm,\al,\bt \vdash [\al\to\bt/a]\jg, x:\al \vdash e\Lto \bt \quad\jimmy{\text{differs from Dunfield's}}\\
\Gm \vdash e_1\ e_2 \To_a \jg &\rto \Gm \vdash e_1\To_{a_1} (\appInfAlg{a_1}{e_2})
\\[2mm]
\Gm \vdash \appInfAlg{\forall a. A}{e} &\rto \Gm,\al \vdash \appInfAlg{[\al/a]A}{e}\\
\Gm \vdash \appInfAlg{A\to C}{e} &\rto \Gm \vdash [C/a]\jg \vdash e \Lto A\\
\Gm[\al] \vdash \appInfAlg{\al}{e} &\rto
	[\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \vdash [\al[2]/a]\jg \vdash e\Lto \al[1])\\
 &\color{magenta} \makebox[0pt]{\qquad or}\phantom{{}\rto{}}
 	[\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \vdash \appInfAlg{\al[1]\to\al[2]}{e}) \quad\text{(1 step equiv.)}
\end{aligned}
\end{gather*}
\caption{Algorithmic Typing}\label{fig:alg}
\end{figure}

>>>>

\begin{figure}
\begin{gather*}
\begin{aligned}
\text{Type variables}\qquad&a, b\\
\text{Existential variables}\qquad&\al, \bt
\\[3mm]
%\text{Types}\qquad&A', B', C' &::=&\quad 1 \mid a \mid \forall x. A' \mid A'\to B'\\
\text{Mono-types}\qquad&\tau &::=&\quad 1 \mid a \mid \tau_1\to \tau_2\\
%\text{Context}\qquad&\Psi &::=&\quad \cdot \mid \Psi, a \mid \Psi, x:A' \color{red} \\[3mm]
% \text{Declarative Judgements}\qquad&\jo &::=&\quad \cdot \mid \P\vdash A' \le B' \jc \jo\\
% &&&\quad \mid \P\vdash e\Leftarrow A' \jc \jo \mid \P\vdash e\To A' \jc \jo \mid \P\vdash A' \bullet e\TTo C' \jc \jo\\
% \noalign{\jimmy{\text{What do the markers do: $\G \vdash,a \vdash,b\vdash \j_1;\j_2;\j_3$ means $\G,a,b\vdash \j_1 \land \G,a\vdash \j_1,\j_2 \land \G\vdash \j_3$}}} \\[3mm]
\text{Algorithmic types}\qquad&A, B, C &::=&\quad 1 \mid a \mid \al \mid \forall a. A \mid A\to B
\\[3mm]
\text{Terms}\qquad&e&::=&\quad x \mid () \mid \lam{e} \mid e_1~e_2 \mid (e:A)
\\[3mm]
\text{Algorithmic judgement chain}\qquad&\jg &::=&\quad A \le B \mid e\Lto A \mid e\To_{a} \jg \mid \appInfAlg{A}{e}
\\[3mm]
\text{Algorithmic worklist}\qquad&\Gm &::=&\quad \cdot \mid \Gm, a \mid \Gm, \al \mid \Gm, x: A \mid \Gm \vdash \jg\\
\text{Declarative worklist}\qquad&\Om &::=&\quad \cdot \mid \Gm, a \mid \Gm, x: A \mid \Gm \vdash \jg
\end{aligned}
\end{gather*}
\caption{Algorithmic Syntax}\label{fig:alg_syntax}
\end{figure}

@@

\paragraph{Judgments and judgment chains} Judgment is a core component during algorithmic type-checking.
There are four kinds of judgments in our system: subtyping, checking, inference and application inference.
Subtyping and checking are relatively simple, since the result could be either success or fail, and no values will be returned.
For inference and application inference, they would return some type that leads to further computation, so we use continuation-passing-style encodings for them.
For example, a judgment chain $e \To_a (a \le B)$ contains two judgments: first we want to infer the type of the expression $e$, and then check if that type is a subtype of $B$.
In the rest of the paper, we usually refer to ``judgment chains'' by ``judgments''.

\paragraph{Worklist Judgments} Our algorithm has a non-standard form.
We combine the traditional contexts and judgment(s) into a single worklist.
The worklist collects variable bindings and judgments in \emph{order}, which represents their scopings:
each variable is visible only after it is declared.
For example, $[a,x:a,x \Lto a]$ is a valid worklist, but $[\textcolor{red}{x} \Lto a, x:\textcolor{red}{a},a]$ is not (red-colored symbols are out-of-scope).

@@

\begin{figure}[t]
\framebox{$\Gm\vdash\jg$} Well-formed algorithmic judgment
\begin{gather*}
\inferrule*[right=$\mathtt{wf{\le}}$]
{\Gm\vdash A \\ \Gm\vdash B}
{\Gm\vdash A\le B}
\qquad
\inferrule*[right=$\mathtt{wf{\Lto}}$]
{\Gm\vdash e \\ \Gm\vdash A}
{\Gm\vdash e \Lto A}
\\
\inferrule*[right=$\mathtt{wf{\To}}$]
{\Gm\vdash e \\ \Gm, a\vdash \jg}
{\Gm\vdash e \To_a \jg}
\qquad
\inferrule*[right=$\mathtt{wf{\TTo}}$]
{\Gm\vdash A \\ \Gm, a\vdash \jg \\ \Gm\vdash e}
{\Gm\vdash \appInfAlg{A}{e}}
\end{gather*}

\framebox{$\text{wf }\Gm$} Well-formed algorithmic worklist
\begin{gather*}
\inferrule*[right=$\mathtt{wf\cdot}$]
{~}{\text{wf }\cdot}
\qquad
\inferrule*[right=$\mathtt{wf_a}$]
{\text{wf }\Gm}
{\text{wf }\Gm, a}
\qquad
\inferrule*[right=$\mathtt{wf_{\al}}$]
{\text{wf }\Gm}
{\text{wf }\Gm, \al}
\qquad
\inferrule*[right=$\mathtt{wf_{of}}$]
{\text{wf }\Gm \\ \Gm\vdash A}
{\text{wf }\Gm, x:A}
\qquad
\inferrule*[right=$\mathtt{wf_{\jg}}$]
{\text{wf }\Gm \\ \Gm\vdash\jg}
{\text{wf }\Gm \vdash\jg}
\end{gather*}
\caption{Algorithmic Well-Formedness}\label{fig:alg_wf}
\end{figure}

@@

% TODO rules too long
\begin{figure}[t]
\framebox{$\Gm\rto \Gm'$} $\Gm$ reduces to $\Gm'$.
%\jimmy{Whenever there is a change on judgment count, marker should be properly added/removed. Decidability: all valid judgment chain should be reduced to nil.}
\begin{gather*}
\begin{aligned}
\Gm, a &\rto \Gm \qquad\qquad
\Gm, \al \rto \Gm \qquad\qquad
\Gm, x:A \rto \Gm
\\[2mm]
\Gm \vdash 1\le 1 &\rto \Gm\\
\Gm \vdash a\le a &\rto \Gm\\
\Gm \vdash \al\le \al &\rto \Gm
\\[2mm]
\Gm \vdash A_1\to A_2 \le B_1\to B_2 &\rto \Gm \vdash A_2 \le B_2 \vdash B_1\le A_1\\
\Gm \vdash \forall a. A\le B &\rto \Gm,\al \vdash [\al/a]A\le B \quad\text{\jimmy{No more exvar leaks}}\\
\Gm \vdash A\le \forall b. B &\rto \Gm,b \vdash A\le B \quad\text{\jimmy{No more typevar leaks}}
\\[2mm]
%\\
%\text{Let } \color{red}\Gm[\al \toto B // G_M] &:= \Gm_L, G_M, [B/\al]\Gm_R, \text{ when } \Gm[\al] = \Gm_L,\al,\Gm_R \ (G_M\text{ defaults to }\cdot)\\
\Gm[\al] \vdash \al \le A\to B &\rto [\al[1]\to\al[2]/\al] (\Gm[\al[1], \al[2]] \vdash \al[1]\to \al[2] \le A \to B)\\
 &\qquad \text{when }\al\notin FV(A)\cup FV(B)\\
\Gm[\al] \vdash A\to B \le \al &\rto [\al[1]\to \al[2]/\al] (\Gm[\al[1], \al[2]] \vdash A \to B \le \al[1]\to \al[2])\\
 &\qquad \text{when }\al\notin FV(A)\cup FV(B)
 \\[2mm]
\Gm[\al][\bt] \vdash \al \le \bt &\rto [\al/\bt](\Gm[\al][])\\
\Gm[\al][\bt] \vdash \bt \le \al &\rto [\al/\bt](\Gm[\al][])\\
\Gm[a][\bt] \vdash a \le \bt &\rto [a/\bt](\Gm[a][])\\
\Gm[a][\bt] \vdash \bt \le a &\rto [a/\bt](\Gm[a][])\\
\Gm[\bt] \vdash 1 \le \bt &\rto [1/\bt](\Gm[])\\
\Gm[\bt] \vdash \bt \le 1 &\rto [1/\bt](\Gm[])
\\[2mm]
\Gm \vdash e \Lto B &\rto \Gm \vdash e\To_a a\le B\\
\Gm \vdash () \Lto 1 &\rto \Gm\\
\Gm \vdash e\Lto \forall a. A &\rto \Gm,a \vdash e\Lto A\\
\Gm \vdash \lambda x. e \Lto A\to B &\rto \Gm, x:A  \vdash e \Lto B\\
\Gm[\al] \vdash \lambda x. e \Lto \al &\rto [\al[1]\to \al[2] / \al](\Gm[\al[1],\al[2]], x:\al[1] \vdash e \Lto \al[2]) \quad\text{\jimmy{Additional}}
\\[2mm]
\Gm \vdash x\To_a \jg &\rto \Gm \vdash [A/a] \jg \qquad (x:A)\in \Gm\\
\Gm \vdash (e:A)\To_a \jg &\rto \Gm \vdash [A/a]\jg \vdash e \Lto A\\
\Gm \vdash ()\To_a \jg &\rto \Gm \vdash [1/a]\jg \qquad\jimmy{\downarrow\text{Precise control on exvars}}\\
\Gm \vdash \lambda x. e \To_a \jg &\rto
	\Gm,\al,\bt \vdash [\al\to\bt/a]\jg, x:\al \vdash e\Lto \bt \quad\jimmy{\text{differs}}\\
\Gm \vdash e_1\ e_2 \To_a \jg &\rto \Gm \vdash e_1\To_{a_1} (\appInfAlg{a_1}{e_2})
\\[2mm]
\Gm \vdash \appInfAlg{\forall a. A}{e} &\rto \Gm,\al \vdash \appInfAlg{[\al/a]A}{e}\\
\Gm \vdash \appInfAlg{A\to C}{e} &\rto \Gm \vdash [C/a]\jg \vdash e \Lto A\\
\Gm[\al] \vdash \appInfAlg{\al}{e} &\rto
	[\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \vdash [\al[2]/a]\jg \vdash e\Lto \al[1])\\
 &\color{magenta} \makebox[0pt]{\qquad or}\phantom{{}\rto{}}
 	[\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \vdash \appInfAlg{\al[1]\to\al[2]}{e})
\end{aligned}
\end{gather*}
\caption{Algorithmic Typing}\label{fig:alg}
\end{figure}
